---
title: Weather and Daily City Bike Usage
author: Joshua Krusell
institute: "Replication code available at https://github.com/jsks/RAV"
date: 2019-11-05
header-includes:
    - \usepackage{amsmath,amssymb}
output:
    beamer_presentation:
        latex_engine: xelatex
        theme: metropolis
        fonttheme: structurebold
---

```{r include = F, results = "hide"}
library(bayesplot)
library(dplyr)
library(ggplot2)
library(rstan)

theme_set(theme_bw() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()))

df <- readRDS("data/dataset.rds")
```

## Weather Data

```{r echo = F}
par(mfrow = c(3, 1))
with(df, plot(datestr, temp))
with(df, plot(datestr, wind))
with(df, plot(datestr, precipitation))
```

## Bicycle Usage

```{r echo = F}
sub.df <- filter(df, !is.na(count), !is.na(year)) %>%
    mutate(monthDay = format(datestr, "%m-%d"))
breaks <- paste0(formatC(4:11, width = 2, flag = 0), "-01")

ggplot(sub.df, aes(monthDay, count, group = 1)) +
    geom_point(na.rm = T) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_grid(year ~ .) +
    ylab("Ridership") +
    theme_bw() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    scale_x_discrete(breaks = breaks)
```

## Bicycle Usage

```{r echo = F}
par(mfrow = c(2, 1))
with(sub.df, boxplot(count ~ month))
with(sub.df, boxplot(count ~ weekday))
```

## Scatterplot Matrix

```{r echo = F}
select(sub.df, ridership = count, temp, wind, precipitation) %>%
    data.matrix %>%
    pairs
```

## Model

Let:

- $y \in \mathbb{N}^n$ be the daily ridership count
- $X \in \mathbb{R}^{n \times m}$ the matrix of predictors
- $\zeta^k \in \mathbb{R}^{J_k}$ the set of partially pooled random
  intercepts for $K$ grouping factors
\begin{align*}
    y_i & \sim \mathsf{NegBinomial}(\mu_i, \phi) \\
    \mathsf{log}(\mu_i) & = \alpha + X_i \beta + \sum_{k=1}^K \zeta^k_{j_i} \\
    \mathop{\mathbb{E}}[Y] & = \mu \mbox{ and }
        \mathsf{Var}[Y] = \mu + \frac{\mu^2}{\phi}
\end{align*}

## Priors

Weakly informative priors:
\begin{align*}
    \alpha & \sim \mathsf{Normal}(0, 10) \\
    \beta & \sim \mathsf{Normal}(0, 5) \\
    \zeta^k & \sim \mathsf{Normal}(0, \sigma^k) \\
    \sigma^k & \sim \mathsf{HalfCauchy}(0, 1) \\
    \phi & \sim \mathsf{HalfCauchy}(0, 1)
\end{align*}

## Results

```{r echo = F}
coefficients <- readRDS("posterior/coef.rds")
mcmc_areas(coefficients[, -1])
```

## Posterior Predictive Checking

```{r echo = F}
df <- filter(df, count > 500)
y_rep <- readRDS("posterior/y_rep.rds")

ppc_dens_overlay(df$count, y_rep[1:50, ])
```

## Posterior Predictive Error

```{r echo = F}
v <- sapply(1:ncol(y_rep), function(i) (df$count[i] - y_rep[, i]) ^ 2) %>%
    rowMeans %>%
    sqrt %>%
    quantile(., probs = c(0.1, 0.9))  %>%
    signif
```

RMSE: `r sprintf("%.2f - %.2f", v[1], v[2])`

```{r echo = F}
ll <- lapply(1:ncol(y_rep), function(i) {
    quantile(df$count[i] - y_rep[, i], probs = c(0.1, 0.5, 0.9))
}) %>% do.call(rbind, .)

errors.df <- as.data.frame(ll) %>%
    rename(codelow = `10%`,
           median = `50%`,
           codehigh = `90%`) %>%
    mutate(y = df$count)

ggplot(errors.df, aes(y, median, group = 1)) +
    geom_errorbar(aes(y, ymin = codelow, ymax = codehigh),
                  colour = "lightblue", width = 0) +
    ylab("Observed - Predicted") +
    xlab("Observed Ridership")

```

## Posterior Predictions

```{r echo = F}
df <- filter(df, !is.na(count), count > 500)

pred.df <- apply(y_rep, 2, quantile, probs = c(0.1, 0.5, 0.9)) %>%
    t %>%
    as.data.frame %>%
    rename(codelow = `10%`,
           median = `50%`,
           codehigh = `90%`) %>%
    mutate(monthDay = format(df$datestr, "%m-%d"),
           y = df$count,
           year = df$year)

ggplot(pred.df, aes(monthDay, y, group = 1)) +
    geom_line() +
    geom_ribbon(aes(ymin = codelow, ymax = codehigh),
                alpha = 0.5, fill = "lightblue") +
    facet_grid(year ~ .) +
    scale_x_discrete(breaks = breaks)
```


## Marginal Effect - Temperature

```{r echo = F}
v <- seq(min(df$temp), max(df$temp), by = 0.1)
m <- lapply(v, function(x) {
        exp(coefficients[, "intercept"] + coefficients[, "temp"] * x) %>%
            quantile(probs = c(0.1, 0.5, 0.9))
}) %>% do.call(rbind, .)

marginal.df <- as.data.frame(m) %>%
    rename(codelow = `10%`,
           median = `50%`,
           codehigh = `90%`) %>%
    mutate(temp = v)

ggplot(marginal.df, aes(temp, median, group = 1)) +
    geom_line() +
    geom_ribbon(aes(ymin = codelow, ymax = codehigh),
                alpha = 0.5, fill = "lightblue") +
    ylab("Ridership") +
    xlab("Temperature")
```

## Marginal Effect - Precipitation

```{r echo = F}
v <- seq(min(df$precipitation), max(df$precipitation), by = 0.1)
m <- lapply(v, function(x) {
        exp(coefficients[, "intercept"] + coefficients[, "precipitation"] * x) %>%
            quantile(probs = c(0.1, 0.5, 0.9))
}) %>% do.call(rbind, .)

marginal.df <- as.data.frame(m) %>%
    rename(codelow = `10%`,
           median = `50%`,
           codehigh = `90%`) %>%
    mutate(precipitation = v)

ggplot(marginal.df, aes(precipitation, median, group = 1)) +
    geom_line() +
    geom_ribbon(aes(ymin = codelow, ymax = codehigh),
                alpha = 0.5, fill = "lightblue") +
    ylab("Ridership") +
    xlab("Precipitation")

```

## Limitations

- Time endogeneity
- Weather aggregation
- Holidays
- Non-parametric functional terms: gaussian processes!
