---
title: "Title"
author: Joshua Krusell
date: 2019-11-04
header-includes:
    - \usepackage{amsmath,amssymb}
output:
    beamer_presentation:
        latex_engine: xelatex
        theme: metropolis
        fonttheme: structurebold
---

```{r include = F, results = "hide"}
library(dplyr)
library(ggplot2)
library(rstan)

theme_set(theme_bw() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()))

df <- readRDS("data/dataset.rds")
#fit <- readRDS("data/fit.rds")
```

## Weather Data

```{r echo = F}
par(mfrow = c(3, 1))
with(df, plot(datestr, temp))
with(df, plot(datestr, wind))
with(df, plot(datestr, precipitation))
```

## Bicycle Usage

```{r echo = F}
sub.df <- filter(df, !is.na(count), !is.na(year)) %>%
    mutate(monthDay = format(datestr, "%m-%d"))
breaks <- paste0(formatC(4:11, width = 2, flag = 0), "-01")

ggplot(sub.df, aes(monthDay, count, group = 1)) +
    geom_point(na.rm = T) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_grid(year ~ .) +
    ylab("Ridership") +
    theme_bw() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    scale_x_discrete(breaks = breaks)
```

## Bicycle Usage

```{r echo = F}
par(mfrow = c(2, 1))
with(sub.df, boxplot(count ~ month))
with(sub.df, boxplot(count ~ weekday))
```

## Scatterplot Matrix

```{r echo = F}
select(sub.df, ridership = count, temp, wind, precipitation) %>%
    data.matrix %>%
    pairs
```

## Model

Let:

- $y \in \mathbb{N}^n$ be the daily ridership count
- $X \in \mathbb{R}^{n \times m}$ the matrix of predictors
- $\zeta^k \in \mathbb{R}^{J_k}$ the set of partially pooled random
  intercepts for $K$ grouping factors
\begin{align*}
    y_i & \sim \mathsf{NegBinomial}(\mu_i, \phi) \\
    \mathsf{log}(\mu_i) & = \alpha + X_i \beta + \sum_{k=1}^K \zeta^k_{j_i} \\
    \mathop{\mathbb{E}}[Y] & = \mu \mbox{ and }
        \mathsf{Var}[Y] = \mu + \frac{\mu^2}{\phi}
\end{align*}

## Priors

Weakly informative priors:
\begin{align*}
    \alpha & \sim \mathsf{Normal}(0, 10) \\
    \beta & \sim \mathsf{Normal}(0, 5) \\
    \zeta^k & \sim \mathsf{Normal}(0, \sigma^k) \\
    \sigma^k & \sim \mathsf{HalfCauchy}(0, 1) \\
    \phi & \sim \mathsf{HalfCauchy}(0, 1)
\end{align*}
